# Session 4: Data visualisation

## Presenting your results in a table

And use functions from another package to display the information in a more user-friendly table.

The `gt` package provides a very flexible interface for building tables from your data.

```{r}
pacman::p_load(gt)
```

The documentation describing the functions can be found [here](https://gt.rstudio.com/articles/case-study-gtcars.html).

Below is an example using the dataset we have built.

```{r}
northern_africa_cases_country_table <- northern_africa_cases_country %>% 
  gt() %>% 
   tab_header(
    title = md("COVID-19 in Northern Africa")
  ) %>% 
  cols_label(
    country = "Country",
    total_covid_cases = "N",
    percentage = "% of total cases in Northern Africa"
  ) %>% 
    tab_spanner(
    label = "Confirmed cases",
    columns = c(total_covid_cases,percentage)
  ) %>% 
    fmt_number(
    columns = total_covid_cases,
    decimals=0,
    use_seps = TRUE
  ) %>% 
   cols_align(
    align = "center",
    columns = c(total_covid_cases, percentage)
  ) 
northern_africa_cases_country_table
```

`gt` has many options for customising tables. To demonstrate this, we will build a table to show when each country in Africa recorded its first COVID-19 case. This example uses some of the techniques demonstrated in [this article](https://www.liamdbailey.com/post/making-beautiful-tables-with-gt/).

```{r}
first_cases_africa <- africa_covid_cases_long %>% 
  select(date=date_format,region=AFRICAN_REGION, country=COUNTRY_NAME, cases) %>% 
  group_by(region,country) %>% 
  filter(cases>0) %>% 
  filter(date == min(date, na.rm=TRUE)) %>% 
  ungroup()

first_cases_africa_table <- first_cases_africa %>% 
  select(region,country,date) %>% 
  group_by(region) %>% 
  arrange(date) %>% 
gt() %>% 
  tab_header(
    title = md("When did countries in Africa record their first case of COVID-19?")
  ) %>% 
  fmt_date(
    columns = date,
    date_style = 4
  ) %>% 
  opt_all_caps() %>% 
  #Use the Chivo font
  #Note the great 'google_font' function in 'gt' that removes the need to pre-load fonts
  opt_table_font(
    font = list(
      google_font("Chivo"),
      default_fonts()
    )
  ) %>%
  cols_label(
    country = "Country",
    date = "Date"
  )  %>% 
  cols_align(
    align = "center",
    columns = c(country, date)
  ) %>% 
  tab_options(
    column_labels.border.top.width = px(3),
    column_labels.border.top.color = "transparent",
    table.border.top.color = "transparent",
    table.border.bottom.color = "transparent",
    data_row.padding = px(3),
    source_notes.font.size = 12,
    heading.align = "left",
    #Adjust grouped rows to make them stand out
    row_group.background.color = "grey") %>% 
  tab_source_note(source_note = "Data: Compiled from national governments and WHO by Humanitarian Emergency Response Africa (HERA)")

first_cases_africa_table
```

## Visualising data using ggplot

One of the key strengths of R is visualising data. There are many packages which have functions you can use to make graphs, tables, maps...the list is endless!

The first package of functions we will use for visualising data is another core `tidyverse` package called `ggplot2.` This is commonly referred to as `ggplot`

We have already loaded the package when we ran `library(tidyverse)`

You can also choose to only load the ggplot2 package by typing `library(ggplot2)`

```{r}
library(ggplot2)
```

The Epidemiologist R handbook has 2 sections focused on `ggplot`

1.  [ggplot basics](https://epirhandbook.com/ggplot-basics.html)

2.  [ggplot tips](https://epirhandbook.com/ggplot-tips.html)

These sections contain very helpful explanations of many of the functions available with ggplot. There are also a number of excellent references for every type of graph you want to make.

We will walk through some common examples to teach some of the most common approaches

### Epicurves

Firstly, we will produce epicurves to describe the distribution of COVID-19 cases (y axis) over time (x axis).

Make a graph of confirmed COVID-19 cases in Northern Africa

```{r}
ggplot(northern_africa, aes(x=date,y=cases)) +
  geom_line()
```

This command has generated a line graph of confirmed COVID-19 cases for countries in Northern Africa.

From earlier steps, we know that the dataset `northern_africa` contains data from multiple countries: \`r unique(northern_africa\$country'

We can add more information to the ggplot command to draw separate lines for each country

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_line()
```

To make the graph more presentable, we can add more options to the ggplot command

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_line() +
  labs(x='Date', y='Total cases', color='Country') + #label axes
  theme(legend.position='top') + #place legend at top of graph
  scale_x_date(date_breaks = '2 months', #set x axis to have 2 month breaks
               date_minor_breaks = '1 month', #set x axis to have 1 month breaks
               date_labels = '%d-%m-%y') #change label for x axis

```

More information on plotting time-series data using ggplot can be found [here](https://www.r-graph-gallery.com/279-plotting-time-series-with-ggplot2.html).

It is still difficult to see the data for each country. There is a helpful command called `facet_wrap` to fix this and allow us to show multiple epicurves by country.

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_col() +
  labs(x='Date', y='Total cases') + #label axes
  theme(legend.position='none') + #remove legend by setting position to 'none'
  scale_x_date(date_breaks = '4 months', #set x axis to have 2 month breaks
               date_minor_breaks = '2 months', #set x axis to have 1 month breaks
               date_labels = '%m-%Y') + #change label for x axis
  facet_wrap(~country) # this will create a separate graph for each country
```

### Visualising moving average data

In a previous section, we added indicators for the rolling average and rolling sum of cases. These indicators can be helpful for identifying trends over time.

```{r}
moroocco_covid_cases_graph <- morocco_covid_cases_mean %>% 
  ggplot() +
  geom_col(aes(x=date, y=cases, color=country)) +
  geom_line(aes(x=date, y=cases_7day_mean)) +
  labs(x='Month-Year', 
       y='Total cases', 
       title='Cases and 7-day average (black line)') +
  theme(legend.position='none') +
  scale_x_date(date_breaks = '2 months', 
               date_minor_breaks = '1 month',
               date_labels = '%m-%y') 

moroocco_covid_cases_graph
```

This chart shows the total number of COVID-19 cases for each day in Morocco between January 1, 2020 and May 3, 2021. The red bars show the reported case numbers for each day while the black line show the 7-day average of cases. We can see that there are several dates with substantially higher numbers of cases compared to the neighbouring dates. This could be due to increased testing on specific days but it is more likely due to delays in reporting leading to a backlog of cases reported on specific days. The black line "smooths" out these differences, allowing us to see the overall trend.

## Useful resources

[Data Visualization: A practical introduction](https://socviz.co/)
