# Session 4: Presenting your data

## Presenting your results in a table

And use functions from another package to display the information in a more user-friendly table.

The `gt` package provides a very flexible interface for building tables from your data.

```{r}
pacman::p_load(gt)
```

The documentation describing the functions can be found [here](https://gt.rstudio.com/articles/case-study-gtcars.html).

Below is an example using the dataset we have built.

```{r}
northern_africa_cases_country_table <- northern_africa_cases_country %>% 
  gt() %>% 
   tab_header(
    title = md("COVID-19 in Northern Africa")
  ) %>% 
  cols_label(
    country = "Country",
    total_covid_cases = "N",
    percentage = "% of total cases in Northern Africa"
  ) %>% 
    tab_spanner(
    label = "Confirmed cases",
    columns = c(total_covid_cases,percentage)
  ) %>% 
    fmt_number(
    columns = total_covid_cases,
    decimals=0,
    use_seps = TRUE
  ) %>% 
   cols_align(
    align = "center",
    columns = c(total_covid_cases, percentage)
  ) 
northern_africa_cases_country_table
```

`gt` has many options for customising tables. To demonstrate this, we will build a table to show when each country in Africa recorded its first COVID-19 case. This example uses some of the techniques demonstrated in [this article](https://www.liamdbailey.com/post/making-beautiful-tables-with-gt/).

```{r}
first_cases_africa <- africa_covid_cases_long %>% 
  select(date=date_format,region=AFRICAN_REGION, country=COUNTRY_NAME, cases) %>% 
  group_by(region,country) %>% 
  filter(cases>0) %>% 
  filter(date == min(date, na.rm=TRUE)) %>% 
  ungroup()

first_cases_africa_table <- first_cases_africa %>% 
  select(region,country,date) %>% 
  group_by(region) %>% 
  arrange(date) %>% 
gt() %>% 
  tab_header(
    title = md("When did countries in Africa record their first case of COVID-19?")
  ) %>% 
  fmt_date(
    columns = date,
    date_style = 4
  ) %>% 
  opt_all_caps() %>% 
  #Use the Chivo font
  #Note the great 'google_font' function in 'gt' that removes the need to pre-load fonts
  opt_table_font(
    font = list(
      google_font("Chivo"),
      default_fonts()
    )
  ) %>%
  cols_label(
    country = "Country",
    date = "Date"
  )  %>% 
  cols_align(
    align = "center",
    columns = c(country, date)
  ) %>% 
  tab_options(
    column_labels.border.top.width = px(3),
    column_labels.border.top.color = "transparent",
    table.border.top.color = "transparent",
    table.border.bottom.color = "transparent",
    data_row.padding = px(3),
    source_notes.font.size = 12,
    heading.align = "left",
    #Adjust grouped rows to make them stand out
    row_group.background.color = "grey") %>% 
  tab_source_note(source_note = "Data: Compiled from national governments and WHO by Humanitarian Emergency Response Africa (HERA)")

first_cases_africa_table
```

## Visualising data using ggplot

One of the key strengths of R is visualising data. There are many packages which have functions you can use to make graphs, tables, maps...the list is endless!

The first package of functions we will use for visualising data is another core `tidyverse` package called `ggplot2.` This is commonly referred to as `ggplot`

We have already loaded the package when we ran `library(tidyverse)`

You can also choose to only load the ggplot2 package by typing `library(ggplot2)`

```{r}
library(ggplot2)
```

The Epidemiologist R handbook has 2 sections focused on `ggplot`

1.  [ggplot basics](https://epirhandbook.com/ggplot-basics.html)

2.  [ggplot tips](https://epirhandbook.com/ggplot-tips.html)

These sections contain very helpful explanations of many of the functions available with ggplot. There are also a number of excellent references for every type of graph you want to make.

We will walk through some common examples to teach some of the most common approaches

### Epicurves

Firstly, we will produce epicurves to describe the distribution of COVID-19 cases (y axis) over time (x axis).


Make a graph of confirmed COVID-19 cases in Northern Africa

```{r}
ggplot(northern_africa, aes(x=date,y=cases)) +
  geom_line()
```

This command has generated a line graph of confirmed COVID-19 cases for countries in Northern Africa.

From earlier steps, we know that the dataset `northern_africa` contains data from multiple countries: \`r unique(northern_africa\$country'

We can add more information to the ggplot command to draw separate lines for each country

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_line()
```

To make the graph more presentable, we can add more options to the ggplot command

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_line() +
  labs(x='Date', y='Total cases', color='Country') + #label axes
  theme(legend.position='top') + #place legend at top of graph
  scale_x_date(date_breaks = '2 months', #set x axis to have 2 month breaks
               date_minor_breaks = '1 month', #set x axis to have 1 month breaks
               date_labels = '%d-%m-%y') #change label for x axis

```

More information on plotting time-series data using ggplot can be found [here](https://www.r-graph-gallery.com/279-plotting-time-series-with-ggplot2.html).

It is still difficult to see the data for each country. There is a helpful command called `facet_wrap` to
fix this and allow us to show multiple epicurves by country.

```{r}
ggplot(northern_africa, aes(x=date,y=cases, color=country)) +
  geom_col() +
  labs(x='Date', y='Total cases') + #label axes
  theme(legend.position='none') + #remove legend by setting position to 'none'
  scale_x_date(date_breaks = '4 months', #set x axis to have 2 month breaks
               date_minor_breaks = '2 months', #set x axis to have 1 month breaks
               date_labels = '%m-%Y') + #change label for x axis
  facet_wrap(~country) # this will create a separate graph for each country
```

### Visualising moving average data

In a previous section, we added indicators for the rolling average and rolling sum of cases. These indicators can be helpful for identifying trends over time.

```{r}
moroocco_covid_cases_graph <- morocco_covid_cases_mean %>% 
  ggplot() +
  geom_col(aes(x=date, y=cases, color=country)) +
  geom_line(aes(x=date, y=cases_7day_mean)) +
  labs(x='Month-Year', 
       y='Total cases', 
       title='Cases and 7-day average (black line)') +
  theme(legend.position='none') +
  scale_x_date(date_breaks = '2 months', 
               date_minor_breaks = '1 month',
               date_labels = '%m-%y') 

moroocco_covid_cases_graph
```

This chart shows the total number of COVID-19 cases for each day in Morocco between January 1, 2020 and May 3, 2021. The red bars show the reported case numbers for each day while the black line show the 7-day average of cases. We can see that there are several dates with substantially higher numbers of cases compared to the neighbouring dates. This could be due to increased testing on specific days but it is more likely due to delays in reporting leading to a backlog of cases reported on specific days. The black line "smooths" out these differences, allowing us to see the overall trend.

## Geographic Information Systems (GIS)

There are many R packages available for applying method from Geographic Information Systems (GIS) to your data.

The Epidemiologist R handbook has an extensive section - [28: GIS basics](https://epirhandbook.com/gis-basics.html)

Some of the examples touch on more advanced Spatial Epidemiology techniques which can be used to derive insights from your data frame. For the purpose of this training, we will focus on a sub-section of the GIS basics: 28.8 [Mapping with ggplot2](https://epirhandbook.com/gis-basics.html#mapping-with-ggplot2)

### Shapefiles

There are several key terms which are defined in the Epidemiologist R handbook - [28.2 Key terms](https://epirhandbook.com/gis-basics.html#key-terms-1)

Understanding what these terms mean will help you to understand the main concepts in this section. For example, we will be using a shapefile to tell R the location where we want to build a map. As per section 28.2, a shapefile can be defined as

*"a common data format for storing "vector" spatial data consisting or lines, points, or polygons. A single shapefile is actually a collection of at least three files - .shp, .shx, and .dbf. All of these sub-component files must be present in a given directory (folder) for the shapefile to be readable. These associated files can be compressed into a ZIP folder to be sent via email or download from a website."*

Shapefiles are often made available free of charge by official government bodies or international humanitarian organisations. For this example, we will be using a shapefile of countries in Africa provided by Africa CDC.

### Loading shapefiles

Spatial data can be complex, particularly if the files cover a large area with many complex borders. To simplify the use of spatial data, several packages have been developed to apply the "tidy" data approach to shapefiles. One example is a package called `sf.`

Before using `sf` to read the data and assign it to an object in R, you will need to unzip the shapefile that you have downloaded. When you unzip this file you will see several files with different endings e.g. ".dbf", ".prj", ".shp".

The file we want to import ends in ".shp" but we also want to keep the other files in the folder. The other files contain useful information about the shapefile which GIS programs can use to correctly import the shapefile.

```{r}
pacman::p_load(sf)
africa_shp <- read_sf(here('data', 'country_boundaries', 'Country_Boundaries.shp')) 
```

The shapefile has now been loaded and added to an object called `africa_map`. We can look at this object to understand more about the shapefile

```{r}
africa_shp
# names(africa_map) #to look at the names of the columns in the dataframe
# head(africa_map) #to look at the first few rows of data
```

There are two variables in the dataset - `COUNTRY_NA` & `geometry`. The `geometry` variable is the most important variable when working with sf objects. This contains all of the geographical co-ordinate information which lets R knows how to plot the map.

As we have previously used ggplot2 to visualise data, we will continue to use that package for making simple maps. There are many other packages which can be used to make maps in R: `leaflet`, `tmap`, `mapbox.`

For this example, we want to map the total number of confirmed COVID-19 cases in North Africa. The shapefile we have loaded contains geographical coordinate information for all countries in Africa. We can use methods from previous sections to filter the shapefile to only include countries in Northern Africa.

```{r}
#check which countries are in the northern_africa_cases_country data frame
northern_african_countries <- unique(northern_africa_cases_country$country)
northern_african_countries
#n=6

north_africa_shp <- africa_shp %>% 
  sf::st_simplify() %>% 
  rename(country=COUNTRY_NA) %>% #rename COUNTRY_NA to make it easier to follow the example
  filter(country %in% northern_african_countries) #filter to only include countries in the northern_african_countries data frame
```